### 2019-01-14 Chapters/Chapter2.tex
4th order

At half filling, in the $t << \text{U}$ limit, which correspond to the strong coupling limit, we can apply perturbation theory to obtain the Heisenberg model from the Hubbard Hamiltonian \ref{Hubbard}. First notice that in the case $t = 0$ the ground state corresponds to having all sites single-occupied. There are two possible spin orientations for each sites, so there is a $2^M$ degeneracy. The ground state energy is $0$. The first exited states are obtained by moving one electron from one site to another, thus leaving one empty site and one double-occupied site. The energy of one of these states is $\text{U}$. Now, if we turn the kinetic term on as a perturbation, having two neighbors electrons with antiparallel spin will allow them to hop and reduce the kinetic energy, whereas while being parallel this hopping is not allowed and there is no energy reduction. Therefore, we can see that when turning the kinetic term on antiparallel alignment will be favored.

@@

The sum being over all excited states. Since $\hat{H_t}\ket{\phi}$ is a superposition of states with exactly one double occupied site and one empty site, we see that only excited states with one double occupied site will contribute. These states have energy $E_s = \text{U}$ therefore we have $\bra{\phi'} \hat{H} \ket{\phi} = -\frac{1}{\text{U}}\sum_s \bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi} = \bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi}$. These means that within the subspace generated by the ground states of $\hat{H_0}$, $\hat{H}$ acts as an effective Hamiltonian $\hat{H}_{\text{eff}} = -\frac{\hat{H_t^2}}{\text{U}}$. Now,

@@

>>>>

At half filling and in the $t << \text{U}$ limit, which correspond to the strong coupling limit, we can apply perturbation theory to obtain the Heisenberg model from the Hubbard Hamiltonian \ref{Hubbard}. First notice that in the case $t = 0$ the ground state corresponds to having all sites single-occupied. There are two possible spin orientations for each sites, so there is a $2^M$ degeneracy. The ground state energy is $0$. The first exited states are obtained by moving one electron from one site to another, thus leaving one empty site and one double-occupied site. The energy of one of these states is $\text{U}$. Now, if we turn the kinetic term on as a perturbation, having two neighbors electrons with antiparallel spin will allow them to hop and reduce the kinetic energy, whereas while being parallel this hopping is not allowed and there is no energy reduction. Therefore, we can see that when turning the kinetic term on antiparallel alignment will be favored.

@@

The sum being over all excited states. Since $\hat{H_t}\ket{\phi}$ is a superposition of states with exactly one double occupied site and one empty site, we see that only excited states with one double occupied site will contribute. These states have energy $E_s = \text{U}$ therefore we have $\bra{\phi'} \hat{H} \ket{\phi} = -\frac{1}{\text{U}}\sum_s \bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi} = \bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi}$. This means that within the subspace generated by the ground states of $\hat{H_0}$, $\hat{H}$ acts as an effective Hamiltonian $\hat{H}_{\text{eff}} = -\frac{\hat{H_t^2}}{\text{U}}$. Now,

@@

\begin{section}{Next-nearest neighbor hopping}
This analysis can be easily extended to electronic Hamiltonians including next-nearest neighbor (NNN) hopping term. The simplest example of such a Hamiltonian is:

\begin{equation}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma}t_{1}\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} -\sum_{\langle \langle i,j \rangle \rangle, \sigma} t_{2}\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}

Where $\langle \langle i,j \rangle \rangle$ denotes sum over NNN. If we follow the procedure as before we will obtain an effective spin Hamiltonian:

\begin{equation}
\hat{H}_{\text{eff}} = J_1\sum_{\langle i,j \rangle} \bs{S}_i\bs{S}_j + J_2\sum_{\langle \langle i,j \rangle \rangle}\bs{S}_i\bs{S}_j
\end{equation}

Where $J_1 = \frac{t_1^2}{\text{U}}$ and $J_2 = \frac{t_2^2}{\text{U}}$. This Hamiltonian is known as $J_1-J_2$ Heisenberg model. This model has been extensively studied, and the $J_1-J_2-J_3$ is also of interest \cite{Reuther2011}.

Notice that we can also obtain a $J_1-J_2$ Heisenberg model starting simply from a NN hamiltonian $\hat{H} = -\sum_{\langle i,j \rangle, \sigma}t_{1}\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$. This is because the NNN exchange interaction comes from the possibility of the process $i\sigma \rightarrow j\sigma \rightarrow i\sigma$ being $i,j$ NNN. This process is available when sites $i, j$ have opposite spins (otherwise it would violate the Pauli principle). Now, when $i,j$ are NNN, we can consider the fourth order process $i\sigma \rightarrow k\sigma \rightarrow j\sigma \rightarrow k\sigma \rightarrow i\sigma$, where $k$ is a site between $i$ and $j$. The resulting effective spin Hamiltonian has exchange couplings $J_1 = 2\frac{t_1^2}{\text{U}} - 8 \frac{t_1^4}{\text{U}^3}$ and $J_2 = 2\frac{t^4}{\text{U}}$. A rigorous derivation can be found in \cite{MacDonald1988}.
